---
title: "R for Actuaries - Part X"
subtitle: "Checking and Exploring Data"
author: "Mick Cooney <mcooney@agrippadataconsulting.com>"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: yes
    theme: cerulean
  pdf_document: default
---

```{r set_options, echo=FALSE}
options(width = 80L
       ,warn  = 1)

knitr::opts_chunk$set(tidy  = FALSE
                     ,cache = FALSE
                     ,fig.height =  7
                     ,fig.width  = 11)
```


# Checking and Exploring Data

In this exercise we are going to build claims triangles from transactional claim
data. The claims are from a fictional captive insurer for a large logistics
company with operations all across Europe.

The data has been provided to us in multiple files, with each file sorted
according to country. The data is current as of date 30 June 2017.

Reading data from a file is an important topic, but is not a good place to
start learning: 80% of the time it is routine and boring, the other 20% of the
time it is fiddly and frustrating! We will return to it later.

For now, we have a script that loads some tables into memory.

```{r load_data, echo=TRUE, message=FALSE}
source('load_data_script.R')
```

By the end of the day, we will be able to produce all these tables from the
claims transactions data.



## Initial Data Checks

The first thing to do with a dataset is inspect it: look at the first few rows,
check the number of rows and columns, and check the data types.

We use two tables of data to learn this manipulation: a dataset of aggregated
triangle data, and the claims transaction data.

### Claims Triangle Data

The triangle data is in 'long' form - the development time is its own variable
rather than a column in the data. We will look at how to convert this to
'wide' format in a little while.

```{r inspect_claim_triangles, echo=TRUE}
print(claim_triangles_tbl)

glimpse(claim_triangles_tbl)
```

The data types that have been loaded seem appropriate.


### Claims Transaction Data

The transaction data shows individual movements on each claim.

```{r inspect_claim_transactions, echo=TRUE}
print(claim_transactions_tbl)

glimpse(claim_transactions_tbl)
```

Again, the data appears to have loaded fine.


## Basic Data Manipulation

We first want to learn how to perform basic manipulation - taking subsets of
rows and columns through the `dplyr` package.

### Subsetting Rows and Columns

In the `dplyr` lexicon, we `select()` columns and `filter()` rows. We start by
subsetting only the claims for Ireland (IRL).

```{r filter_irl_rows, echo=TRUE}
irl_data_tbl <- claim_transactions_tbl %>%
    filter(country_code == 'IRL') %>%
    select(country_code, year, claim_id, transaction_date, claim_type, amount)

glimpse(irl_data_tbl)
```

If we put a `-` symbol in front of column names, we exclude those instead.

```{r filter_irl_rows_removal, echo=TRUE}
irl_data_tbl <- claim_transactions_tbl %>% 
    filter(country_code == 'IRL') %>%
    select(-incident_date, -report_date)

glimpse(irl_data_tbl)
```

We can have multiple conditions at once in a `filter()` condition:

```{r filter_irl_2015_data, echo=TRUE}
irl_2015_data_tbl <- claim_transactions_tbl %>% 
    filter(country_code == 'IRL'
          ,year         == 2015
           ) %>%
    select(-incident_date, -report_date)

glimpse(irl_2015_data_tbl)
```


### Sorting Columns

Suppose we want to look at the claims in order of the date of the incident
date. We do this using `arrange()`:

```{r print_data_incident_date, echo=TRUE}
claim_snapshot_tbl %>% 
    arrange(incident_date) %>%
    print()
```

We could also show the rows by reporting date:

```{r print_date_report_date, echo=TRUE}
claim_snapshot_tbl %>% 
    arrange(report_date) %>%
    print()
```


### Creating Columns

We create new columns in our dataset by using `mutate()`:

```{r create_new_column, echo=TRUE}
claim_transactions_tbl %>%
    mutate(reporting_delay_days = as.numeric(report_date - incident_date)) %>%
    print()
```


## Summarising Data

We often want to perform operations on groups of data: for example, calculate
average current claim amounts by year or by country.

We do this using the `group_by()` and `summarise()` functions






