---
title: "R for Actuaries - Part X"
subtitle: "Loading Data"
author: "Mick Cooney <mcooney@agrippadataconsulting.com>"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: yes
    theme: cerulean
  pdf_document: default
---

```{r set_options, echo=FALSE}
options(width = 80L
       ,warn  = 1)

knitr::opts_chunk$set(tidy  = FALSE
                     ,cache = FALSE
                     ,fig.height =  7
                     ,fig.width  = 11)
```


# Loading and Checking Data

In this exercise we are going to build claims triangles from transactional claim
data. The claims are from a fictional captive insurer for a large logistics
company with operations all across Europe.

The data has been provided to us in multiple files, with each file sorted
according to country. The data is current as of date 30 June 2017.

## Loading Data

Looking into the `data` directory we see thirteen files, of the form
`XXX_claims_data.csv` where the XXX is the three-letter country code. Thus, UK
data is in the file `GBR_claims_data.csv` and so on.

### Reading Single Data Files

To get started, we are first going to load in the IRL data and look at it.

```{r load_irl_data, echo=TRUE}
library(tidyverse)

irl_data_tbl <- read_csv('data/IRL_claims_data.csv')

print(irl_data_tbl)
```

It is important to look at your data after you load it!

`glimpse()` is a useful tool here to check a table

```{r glimpse_irl_data, echo=TRUE}
glimpse(irl_data_tbl)
```






