---
title: "Real-world R for Actuaries"
subtitle: "Data Manipulation to Build Claims Triangles"
author: "Mick Cooney <mcooney@agrippadataconsulting.com>"
date: "2019-02-18"
output:
  html_document:
    fig_caption: yes
    theme: spacelab
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 3
    toc_float:
      smooth_scroll: FALSE
  pdf_document: default
---

```{r knit_opts, include = FALSE}
rm(list = ls())

library(tidyverse)
library(scales)
library(purrr)
library(cowplot)
library(feather)



options(width = 80L
       ,warn  = 1)

knitr::opts_chunk$set(tidy  = FALSE
                     ,cache = FALSE
                     ,warning = FALSE
                     ,message = FALSE
                     ,fig.height =  7
                     ,fig.width  = 11)


set.seed(42)
```

# Load Claim Transactions Data

First we load up the transactional data for the claims

```{r load_claim_transactions, echo=TRUE}
data_files <- list.files('data', full.names = TRUE, pattern = 'claims_data\\.csv')

claim_transactions_tbl <- tibble(file_src = data_files) %>%
    mutate(data = map(file_src, read_csv, col_types = cols())) %>%
    unnest() %>%
    dplyr::select(-file_src)

claim_transactions_tbl %>% glimpse()
claim_transactions_tbl %>% print()
```

## Build Quick Year Triangles

Now we want to build a quick claim triangle

```{r build_year_triangles, echo=TRUE}
bad_triangle_tbl <- claim_transactions_tbl %>%
    mutate(claim_year = year
          ,value_year = transaction_date %>% format('%Y') %>% as.numeric) %>%
    dplyr::select(country_code, claim_id, claim_type, claim_year, value_year, amount) %>%
    group_by(country_code, claim_type, claim_year, value_year) %>%
    summarise(total_amount = sum(amount)) %>%
    mutate(dev_year = value_year - claim_year + 1) %>%
    ungroup() %>%
    dplyr::select(country_code, claim_type, claim_year, dev_year, total_amount)


bi_plot_tbl <- bad_triangle_tbl %>% filter(claim_type == 'BI')
pd_plot_tbl <- bad_triangle_tbl %>% filter(claim_type == 'PD')

ggplot(bi_plot_tbl %>% mutate(claim_year = claim_year %>% as.character)) +
    geom_line(aes(x = dev_year, y = total_amount, colour = claim_year)) +
    facet_wrap(~ country_code, scales = 'free_y') +
    scale_y_continuous(labels = scales::comma) +
    xlab('Development Year') +
    ylab('Total Development')

ggplot(pd_plot_tbl %>% mutate(claim_year = claim_year %>% as.character)) +
    geom_line(aes(x = dev_year, y = total_amount, colour = claim_year)) +
    facet_wrap(~ country_code, scales = 'free_y') +
    scale_y_continuous(labels = scales::comma) +
    xlab('Development Year') +
    ylab('Total Development')

```

## Fixing Triangle Calculation

We are not properly rolling forward claims after they are done developing, so
we need to back fill all those.

```{r fill_full_triangle, echo=TRUE}
movements_tbl <- claim_transactions_tbl %>%
    mutate(claim_year = year
          ,trans_year = transaction_date %>% format('%Y') %>% as.numeric
          ,dev_time   = trans_year - claim_year + 1
          ) %>%
    group_by(claim_id, dev_time) %>%
    top_n(1, wt = transaction_date) %>%
    ungroup() %>%
    dplyr::select(country_code, claim_type, claim_year, claim_id, dev_time, amount)



triangle_zero_tbl <- movements_tbl %>%
    dplyr::select(country_code, claim_type, claim_year, claim_id) %>%
    distinct() %>%
    mutate(dev_time = 0, amount = 0)


calc_incr_tbl <- list(triangle_zero_tbl, movements_tbl) %>%
    bind_rows() %>%
    group_by(country_code, claim_type, claim_year, claim_id) %>%
    arrange(dev_time) %>%
    mutate(increment = amount - lag(amount)) %>%
    ungroup() %>%
    arrange(country_code, claim_type, claim_year, claim_id, dev_time) %>%
    replace_na(list(increment = 0))

calc_incr_tbl <- calc_incr_tbl %>%
    dplyr::select(claim_id, claim_type, claim_year, dev_time, increment)


incr_values_tbl <- claim_transactions_tbl %>%
    dplyr::select(country_code, claim_type, claim_year = year, claim_id) %>%
    distinct() %>%
    mutate(data = list(tibble(dev_time = 0:20))) %>%
    unnest() %>%
    left_join(calc_incr_tbl, by = c('claim_id', 'claim_type', 'claim_year', 'dev_time')) %>%
    replace_na(list(increment = 0))


full_dev_tbl <- incr_values_tbl %>%
    group_by(country_code, claim_type, claim_year) %>%
    arrange(dev_time) %>%
    mutate(cuml_amount = cumsum(increment)) %>%
    ungroup() %>%
    arrange(country_code, claim_type, claim_year, claim_id, dev_time)

long_triangle_tbl <- full_dev_tbl %>%
    group_by(country_code, claim_type, claim_year, dev_time) %>%
    summarise(total_cuml = sum(cuml_amount)) %>%
    ungroup() %>%
    filter(claim_year + dev_time <= 2017)


long_bi_tbl <- long_triangle_tbl %>%
    filter(claim_type == 'BI'
          ,claim_year >= 2010
           ) %>%
    mutate(claim_year = claim_year %>% as.character()
          ,total_cuml = total_cuml / 1e6
           )

long_pd_tbl <- long_triangle_tbl %>%
    filter(claim_type == 'PD'
          ,claim_year >= 2010
           ) %>%
    mutate(claim_year = claim_year %>% as.character()
          ,total_cuml = total_cuml / 1e6
           )


ggplot(long_pd_tbl) +
    geom_line(aes(x = dev_time, y = total_cuml, colour = claim_year)) +
    facet_wrap(~ country_code, scales = 'free_y') +
    scale_y_continuous(labels = scales::comma) +
    xlab("Development Time") +
    ylab("Total Amount")
```


## Write to Disk

```{r write_to_disk, echo=TRUE}
long_triangle_tbl %>% write_feather('data/yearly_triangles.feather')
```


# Build Triangles with Incremental Data

The most straightforward method to build the triangles is likely via
incremental values for the claim movements.

The data we have shows the current value for the claim - we need to switch
this data to incrementals first, and then work from that.

Once we have the incremental values, we then calculate the development time.

Rather than our previous approach to calculating the development period for the
increment we instead use the number of days since the incident and divide by
365.25. As we are assuming no development at time 0, we add 1 to each
development year.

```{r build_triangle_incrementals, echo=TRUE}
claim_incrementals_tbl <- claim_transactions_tbl %>%
    group_by(country_code, year, claim_id, claim_type) %>%
    mutate(incremental = diff(c(0, amount))) %>%
    ungroup() %>%
    mutate(dev_days = as.numeric(transaction_date - incident_date)
          ,dev_year = floor(dev_days / 365.25) + 1
           )

claim_incrementals_tbl %>% glimpse()
claim_incrementals_tbl %>% print()
```

We want to ensure that every claim gets an entry for each development period.

To do this we fill out the each development period for each year, and then join
this data on.

To create a cross join of each 'account' year by each development year we use
the function `crossing()` to 

```{r fill_out_dev_periods, echo=TRUE}
acc_year <- claim_incrementals_tbl %>% pull(year)     %>% unique()
dev_year <- claim_incrementals_tbl %>% pull(dev_year) %>% unique()

all_tbl <- claim_incrementals_tbl %>%
    select(country_code, claim_type, year) %>%
    distinct()

full_incr_tbl <- crossing(year     = acc_year
                    ,dev_year = c(0, dev_year)
                     ) %>%
    left_join(all_tbl, by = 'year')
```


```{r calculate_date_development_periods, echo=TRUE}
data_dev_tbl <- claim_incrementals_tbl %>%
    group_by(country_code, year, claim_type, dev_year) %>%
    summarise(total_incr = sum(incremental))


data_dev_tbl %>% glimpse()
data_dev_tbl %>% print()
```


```{r calculate_all_dev_increments, echo=TRUE}
claim_dev_vals_tbl <- full_incr_tbl %>%
    left_join(data_dev_tbl, by = c('country_code', 'year', 'claim_type', 'dev_year')) %>%
    replace_na(list(total_incr = 0)) %>%
    select(country_code, claim_type, year, dev_year, total_incr) %>%
    arrange(country_code, claim_type, year, dev_year)

claim_dev_vals_tbl %>% glimpse()
claim_dev_vals_tbl %>% print()
```


We are now in a position to calculate the cumulative amounts for each year.

```{r calculate_devtable_long, echo=TRUE}
claim_dev_long_tbl <- claim_dev_vals_tbl %>%
    group_by(country_code, claim_type, year) %>%
    arrange(year) %>%
    mutate(dev_amount = cumsum(total_incr)) %>%
    ungroup() %>%
    arrange(country_code, claim_type, year, dev_year)
  
claim_dev_long_tbl %>% glimpse()
claim_dev_long_tbl %>% print()
```

We spread this table to 'wide' format:

```{r spread_devtable_wide, echo=TRUE}
claim_dev_wide_tbl <- claim_dev_long_tbl %>%
    select(-total_incr) %>%
    spread(dev_year, dev_amount)

claim_dev_wide_tbl %>% glimpse()
claim_dev_wide_tbl %>% print()
```



# Split Claims Data

```{r split_claims_data, echo=TRUE}
countries <- claim_transactions_tbl %>% count(country_code) %>% pull(country_code)

for(country in countries) {
    output_file <- paste0('data/', country, '_claims_data.csv')
    
    claim_transactions_tbl %>%
        filter(country_code == country) %>%
        write_csv(path = output_file)
}
```
